\Large\bf FlexibleSUSY --- A spectrum generator generator for supersymmetric models

### 2014-06-03 doc/paper.tex
use as GUT condition TLambda = Alambda * lambda
use proper shell command
typo in the path to the SLHA file
Addressing Dominik's suggestions on section 5 (not finished yet)

In order to create a spectrum generator the SUSY model must be defined
by specifying the gauge group, the field content and mixings as well
as the superpotential and the soft-breaking terms.  From this
information the renormalization group equations, mass matrices,
self-energies, tadpole diagrams and electroweak symmetry breaking
conditions have to be derived.  These expressions must then be
combined in a computer program to allow for a numeric calculation of
the mass spectrum.  In addition most SUSY models require boundary
conditions for the model parameters at a low and a high scale.  For
example in the CMSSM mSUGRA boundary conditions for the soft-breaking
parameters are imposed at the gauge coupling unification scale.
Furthemore, at the $Z$ mass scale the CMSSM is matched to the Standard
Model, which implies conditions for the gauge and Yukawa couplings.
The so defined boundary value problem must be solved numerically until
a set of model parameters has been found consistent with all
user-supplied boundary conditions.

@@

\fs can be downloaded as a gzipped tar file from the
\url{http://flexiblesusy.hepforge.org/}.  To download and install
version 1.0.0 run:
%
\begin{lstlisting}[language=bash]
$ wget https://www.hepforge.org/archive/flexiblesusy/FlexibleSUSY-1.0.0.tar.gz
$ tar -xf FlexibleSUSY-1.0.0.tar.gz
$ cd FlexibleSUSY-1.0.0/
\end{lstlisting}%% $
%
A CMSSM spectrum generator can be created with the following three
commands:
%
\begin{lstlisting}[language=bash]
$ ./createmodel --name=MSSM
$ ./configure --with-models=MSSM
$ make
\end{lstlisting}%% $
%
The first command creates the model directory \code{models/MSSM/}
together with a CMSSM model file.  The \code{configure} script checks
the system requirements and creates the \code{Makefile}.  See
\code{./configure --help} for more options.  Executing \code{make}
will start \mathematica to generate the spectrum generator and compile
it.  The resulting executable can be run like this:
\begin{lstlisting}[language=bash]
$ models/MSSM/run_MSSM.x --slha-input-file=model_fies/MSSM/LesHouches.in.MSSM
\end{lstlisting}%% $
%
When executed, the spectrum generator tries to find a set of \DRbar\
model parameters consistent with all CMSSM boundary conditions for the
parameter point given in the SLHA input file
\code{model_files/MSSM/LesHouches.in.MSSM}.  Afterwards, the pole mass
spectrum and mixing matrices are calculated and written to the command
line in SLHA format \cite{Skands:2003cj,Allanach:2008qq}.  For the
parameter point given in the above example the calculated pole mass
spectrum reads
%
\begin{lstlisting}
Block MASS
   1000021     1.15236966E+03   # Glu
   1000024     3.85774334E+02   # Cha_1
   1000037     6.50460073E+02   # Cha_2
        25     1.14766149E+02   # hh_1
        35     7.06792640E+02   # hh_2
        37     7.11388516E+02   # Hpm_2
        36     7.06523105E+02   # Ah_2
   1000012     3.51856376E+02   # Sv_1
   1000014     3.53042556E+02   # Sv_2
   1000016     3.53046504E+02   # Sv_3
   1000022     2.03889780E+02   # Chi_1
   1000023     3.85760714E+02   # Chi_2
   1000025     6.36544884E+02   # Chi_3
   1000035     6.50133768E+02   # Chi_4
   1000001     9.66656018E+02   # Sd_1
   1000003     1.00983181E+03   # Sd_2
   1000005     1.01651873E+03   # Sd_3
   2000001     1.01653005E+03   # Sd_4
   2000003     1.06089534E+03   # Sd_5
   2000005     1.06090238E+03   # Sd_6
   1000011     2.22570305E+02   # Se_1
   1000013     2.29864536E+02   # Se_2
   1000015     2.29888846E+02   # Se_3
   2000011     3.61946671E+02   # Se_4
   2000013     3.61950866E+02   # Se_5
   2000015     3.63136031E+02   # Se_6
   1000002     8.09787818E+02   # Su_1
   1000004     1.01454197E+03   # Su_2
   1000006     1.01981109E+03   # Su_3
   2000002     1.02015269E+03   # Su_4
   2000004     1.05807759E+03   # Su_5
   2000006     1.05808168E+03   # Su_6
\end{lstlisting}

@@

HighScaleInput = {
   ...
   {T[\[Lambda]], ALambdaInput},
   ...
};
\end{lstlisting}
%
The value of $A_\lambda$ at the GUT scale can be set in the SLHA input
file in \code{EXTPAR} block entry $63$ via
%
\begin{lstlisting}
Block EXTPAR
   61   0.1                  # LambdaInput
   63   -100                 # ALambdaInput
\end{lstlisting}

\section{Setting up a FlexibleSUSY model}
\label{Sec:modfile}
A \fs model is configured via a (\mathematica) model file
\code{FlexibleSUSY.m} in the model directory \code{models/<model>/}.
Many example model files can be found in \code{model_files/}.  The
default model file for the CMSSM looks like this:
%
\begin{lstlisting}[language=Mathematica]
FSModelName = "MSSM";

@@

All non-Standard Model input variables must be specified in the lists
\code{MINPAR} and \code{EXTPAR}.  Theses two variables refer to the
MINPAR and EXTPAR blocks in a SLHA input file \cite{Skands:2003cj}.
The list elements are two-component lists where the first entry is the
SLHA index in the MINPAR or EXTPAR blocks.  The second entry is the
name of the parameter.

Using the the variable \code{EWSBOutputParameters} the user can
specify the model parameters that are output of the electroweak
symmetry breaking conditions.  When doing EWSB, \fs will adjust these
parameters until the EWSB conditons are fulfilled.  In the CMSSM these
are the superpotential parameter \code{\\[Mu]} and its corresponding
soft-breaking parameter \code{B[\\[Mu]]}.

Furthermore, the user has to specify the form of three model
constraints: low-scale, SUSY-scale and high-scale constraints.  In \fs
they are named as \code{LowScale}, \code{SUSYScale} and
\code{HighScale}.  For each constraint there is (i) a scale definition
(named after the constraint), (ii) an initial guess for the scale
(concatenation of the constraint name and \code{FirstGuess}) and (iii)
a list of parameter settings to be applied at the scale (concatenation
of the constraint name and \code{Input}).  Optional a minimum and a
maximum value for the scale can be given (concatenation of the
constraint name and \code{Minimum} or \code{Maximum}, respectively).
The latter can avoid underflows or overflows of the scale value during
the iteration.  This is especially useful in models where the
iteration is very unstable and the value of the scale is very
sensitive to the model parameters.  The meaning of the three
constraints is the following:
%
\begin{itemize}
\item \emph{High-scale constraint:} The high-scale constraint is
  usually the GUT-scale constraint, applied at the scale where the
  gauge couplings $g_1$ and $g_2$ unify.  The high-scale can be
  defined by an equation of the form \code{g1 == g2} or by fixed scale
  value.
%
\item \emph{SUSY-scale constraint:} The SUSY-scale constraint is the
  constraint where \fs imposes the EWSB conditions and calculate the
  pole mass spectrum.  Its scale $M_S$ is usually defined as
  $\sqrt{m_{\tilde{t}_1}m_{\tilde{t}_2}}$.  In the example above it
  has the value $M_S = \sqrt{m_{\tilde{u}_1}m_{\tilde{u}_6}}$, where
  $m_{\tilde{u}_1}$ and $m_{\tilde{u}_6}$ are the \DRbar\ masses of
  the lightest and heaviest up-type squark, respectively.
%
\item \emph{Low-scale constraint:} The low-scale constraint is the
  constraint where the gauge couplings $g_1$, $g_2$, $g_3$ and the
  Yukawa couplings $y_f$ are calculated from given Standard Model
  input $\alpha_{\text{e.m.}}(M_Z)$, $\alpha_{s}(M_Z)$, $m_f^\pole$,
  $M_Z^\pole$, $M_W^\pole$, see
  Sec.~\ref{sec:calculation-of-gauge-couplings}--\ref{sec:calculation-of-yukawa-couplings}.
  This should be done at the $Z$ mass scale $M_Z$.  The gauge
  couplings are calculated automatically and therefore do not need to
  be specified in the \code{LowScaleInput} variable.
\end{itemize}
%
\fs GUT-normalizes all gauge couplings.  Thus, the high-scale
definition takes the simple form \code{g1 == g2}.  Another consequence
is that in the calculation of the VEVs $v_u$ and $v_d$ from $M_Z$ and
$\tan\beta$ at the low-scale the GUT-normalization has to be taken
into account, see the example above.

The list of parameter settings can contain as elements any of the
following:
%
\begin{itemize}
\item Two-component lists of the form \code{\{parameter, value\}},
  which indicates that the \code{parameter} is set to \code{value} at
  the defined scale.  If the \code{value} should be read from the SLHA
  input file, it must be written as \code{LHInput[value]}.  Example:
  %
  \begin{lstlisting}
SUSYScaleInput = {
   {mHd2, m0^2},
   {mHu2, LHInput[mHu2]}
};
  \end{lstlisting}
  %
  In this example the parameter \code{mHd2} is set to the value of
  \code{m0^2}, and \code{mHu2} is set to the value given in the SLHA
  input file in block \code{MSOFTIN}, entry 22.  The SLHA block names
  and keys for the MSSM and NMSSM are defined in SARAH's
  \code{parameters.m} file, see the SARAH manual or
  \cite{Staub:2010jh}.  The Standard Model Yukawa couplings \code{Yu},
  \code{Yd}, \code{Ye} can be calculated automatically from the known
  Standard Model quark and lepton masses, see
  Section~\ref{sec:calculation-of-yukawa-couplings}.  To enable the
  automatic calculation, \code{value} must be set to \code{Automatic},
  see the example above.

@@

>>>>

In order to create a spectrum generator the SUSY model must be defined
by specifying the gauge group, the field content and mixings as well
as the superpotential and the soft-breaking terms.  From this
information the renormalization group equations, mass matrices,
self-energies, tadpole diagrams and electroweak symmetry breaking
(EWSB) conditions have to be derived.  These expressions must then be
combined in a computer program to allow for a numeric calculation of
the mass spectrum.  In addition most SUSY models require boundary
conditions for the model parameters at a low and a high scale.  For
example in the CMSSM mSUGRA boundary conditions for the soft-breaking
parameters are imposed at the gauge coupling unification scale.
Furthemore, at the $Z$ mass scale the CMSSM is matched to the Standard
Model, which implies conditions for the gauge and Yukawa couplings.
The so defined boundary value problem must be solved numerically until
a set of model parameters has been found consistent with all
user-supplied boundary conditions.

@@

\fs can be downloaded as a gzipped tar file from the
\url{http://flexiblesusy.hepforge.org/}.  To download and install
version 1.0.0 run:
%
\begin{lstlisting}[language=bash]
$ wget https://www.hepforge.org/archive/flexiblesusy/FlexibleSUSY-1.0.0.tar.gz
$ tar -xf FlexibleSUSY-1.0.0.tar.gz
$ cd FlexibleSUSY-1.0.0/
\end{lstlisting}%% $
%
A CMSSM spectrum generator can be created with the following three
commands:
%
\begin{lstlisting}[language=bash]
$ ./createmodel --name=MSSM
$ ./configure --with-models=MSSM
$ make
\end{lstlisting}%% $
%
The first command creates the model directory \code{models/MSSM/}
together with a CMSSM model file.  The \code{configure} script checks
the system requirements and creates the \code{Makefile}.  See
\code{./configure --help} for more options.  Executing \code{make}
will start \mathematica to generate the spectrum generator and compile
it.  The resulting executable can be run like this:
\begin{lstlisting}[language=bash]
$ ./models/MSSM/run_MSSM.x \
    --slha-input-file=model_files/MSSM/LesHouches.in.MSSM
\end{lstlisting}%% $
%
When executed, the spectrum generator tries to find a set of \DRbar\
model parameters consistent with all CMSSM boundary conditions for the
parameter point given in the SLHA input file
\code{model_files/MSSM/LesHouches.in.MSSM}.  Afterwards, the pole mass
spectrum and mixing matrices are calculated and written to the command
line in SLHA format \cite{Skands:2003cj,Allanach:2008qq}.  For the
parameter point given in the above example the calculated pole mass
spectrum reads
%
\begin{lstlisting}
Block MASS
   1000021     1.15236966E+03   # Glu
   1000024     3.85774334E+02   # Cha_1
   1000037     6.50460073E+02   # Cha_2
        25     1.14766149E+02   # hh_1
        35     7.06792640E+02   # hh_2
        37     7.11388516E+02   # Hpm_2
        36     7.06523105E+02   # Ah_2
   1000012     3.51856376E+02   # Sv_1
   1000014     3.53042556E+02   # Sv_2
   1000016     3.53046504E+02   # Sv_3
   1000022     2.03889780E+02   # Chi_1
   1000023     3.85760714E+02   # Chi_2
   1000025     6.36544884E+02   # Chi_3
   1000035     6.50133768E+02   # Chi_4
   1000001     9.66656018E+02   # Sd_1
   1000003     1.00983181E+03   # Sd_2
   1000005     1.01651873E+03   # Sd_3
   2000001     1.01653005E+03   # Sd_4
   2000003     1.06089534E+03   # Sd_5
   2000005     1.06090238E+03   # Sd_6
   1000011     2.22570305E+02   # Se_1
   1000013     2.29864536E+02   # Se_2
   1000015     2.29888846E+02   # Se_3
   2000011     3.61946671E+02   # Se_4
   2000013     3.61950866E+02   # Se_5
   2000015     3.63136031E+02   # Se_6
   1000002     8.09787818E+02   # Su_1
   1000004     1.01454197E+03   # Su_2
   1000006     1.01981109E+03   # Su_3
   2000002     1.02015269E+03   # Su_4
   2000004     1.05807759E+03   # Su_5
   2000006     1.05808168E+03   # Su_6
\end{lstlisting}

@@

HighScaleInput = {
   ...
   {T[\[Lambda]], ALambdaInput LambdaInput},
   ...
};
\end{lstlisting}
%
The value of $A_\lambda$ at the GUT scale can be set in the SLHA input
file in \code{EXTPAR} block entry $63$ via
%
\begin{lstlisting}
Block EXTPAR
   61   0.1                  # LambdaInput
   63   -100                 # ALambdaInput
\end{lstlisting}

\section{Setting up a FlexibleSUSY model}
\label{Sec:modfile}

In general a (non-constrained) softly broken SUSY model is defined by
the gauge group, the field content and mixings as well as the
superpotential and the soft-breaking Lagrangian.  In order to create a
spectrum generator for such a SUSY model with \fs, the aforementioned
model properties have to be defined in a SARAH model file.  The SARAH
model file can be put into the \code{sarah/<model>/} directory.  See
the SARAH manual \cite{Staub:2008uz,Staub:2013tta} for a detailed
explanation of how to write such a model file.

The model boundary conditions are defined in the \fs model file
\code{FlexibleSUSY.m}, which has to be located in the model directory
\code{models/<model>/}.  Note, that many example model files can
already be found in \code{model_files/}.  In the following it is
explained how the boundary conditions are defined on the basis of the
CMSSM.  The application to other models is straightforward.  The CMSSM
model file reads:
%
\begin{lstlisting}[language=Mathematica]
FSModelName = "MSSM";

@@

All non-Standard Model input variables must be specified in the lists
\code{MINPAR} and \code{EXTPAR}.  These two variables refer to the
MINPAR and EXTPAR blocks in a SLHA input file \cite{Skands:2003cj}.
The list elements are two-component lists where the first entry is the
SLHA index in the MINPAR or EXTPAR block, respectively.  The second
entry is the name of the input parameter.  In the above example the
input parameters are the universal soft-breaking parameters $m_0$,
$M_{1/2}$, $A_0$ as well as $\tan\beta$ and $\sign\mu$.

Using the the variable \code{EWSBOutputParameters} the user can
specify the model parameters that are output of the electroweak
symmetry breaking consistency conditions.  When imposing the EWSB, \fs
will adjust these parameters until the EWSB conditons are fulfilled.
In the CMSSM example above these are the superpotential parameter
\code{\\[Mu]} and its corresponding soft-breaking parameter
\code{B[\\[Mu]]}.  In the NMSSM the parameters $\kappa$, $|v_s|$ and
$m_s^2$ are usually chosen for this purpose.

Furthermore, the user has to specify three model constraints:
low-scale, SUSY-scale and high-scale.  In \fs they are named as
\code{LowScale}, \code{SUSYScale} and \code{HighScale}.  For each
constraint there is (i) a scale definition (named after the
constraint), (ii) an initial guess for the scale (concatenation of the
constraint name and \code{FirstGuess}) and (iii) a list of parameter
settings to be applied at the scale (concatenation of the constraint
name and \code{Input}).  Optionally a minimum and a maximum value for
the scale can be given (concatenation of the constraint name and
\code{Minimum} or \code{Maximum}, respectively).  The latter can avoid
underflows or overflows of the scale value during the iteration.  This
is especially useful in models where the iteration is very unstable
and the value of the scale is very sensitive to the model parameters.
The meaning of the three constraints is the following:
%
\begin{itemize}
\item \emph{High-scale constraint:} The high-scale constraint is
  usually the GUT-scale constraint, imposed at the scale where the
  gauge couplings $g_1$ and $g_2$ unify.  The high-scale can be
  defined by an equation of the form \code{g1 == g2} or by a fixed
  numerical value.
%
\item \emph{SUSY-scale constraint:} The SUSY-scale is the typical mass
  scale of the SUSY particle spectrum.  At this scale \fs imposes the
  EWSB conditions and calculates the pole mass spectrum.  The
  SUSY-scale, $M_S$, is usually defined as $M_S =
  \sqrt{m_{\tilde{t}_1}m_{\tilde{t}_2}}$.  However, in the example
  above, where sfermion flavour violation is enabled, it has the value
  $M_S = \sqrt{m_{\tilde{u}_1}m_{\tilde{u}_6}}$, where
  $m_{\tilde{u}_1}$ and $m_{\tilde{u}_6}$ are the \DRbar\ masses of
  the lightest and heaviest up-type squark, respectively.
%
\item \emph{Low-scale constraint:} The low-scale constraint is the
  constraint where the SUSY model is matched to the Standard Model.
  This is done by calculating the gauge couplings $g_i$ ($i=1,2,3$)
  and the Yukawa couplings $y_f$ ($f=u,d,e$) from the known Standard
  Model quantities $\alpha_{\text{e.m.}}(M_Z^\pole)$,
  $\alpha_{s}(M_Z^\pole)$, $m_f^\pole$, $M_Z^\pole$, $M_W^\pole$.  The
  details of the calculation are explained in
  sections~\ref{sec:calculation-of-gauge-couplings}--\ref{sec:calculation-of-yukawa-couplings}.
  Currently this scale is fixed to be the $Z$ pole mass scale
  $M_Z^\pole$.  The calculation of the gauge couplings is done
  automatically and is not needed to be specified in the
  \code{LowScaleInput} variable.  The calculation of the Yukawa
  couplings is optional.  If the Yukawa couplings are set to the value
  \code{Automatic}, they are determined automatically from the known
  Standard Model fermion masses $m_f$ as explained in section
  \ref{sec:calculation-of-yukawa-couplings}.
\end{itemize}
%
Note, that \fs GUT-normalizes all gauge couplings.  Thus, the
high-scale definition takes the simple form \code{g1 == g2}.  Another
consequence is that in the calculation of the VEVs $v_u$ and $v_d$
from $M_Z$ and $\tan\beta$ at the low-scale the GUT-normalization has
to be taken into account, see the example above.

The list of parameter settings for imposing a constraint can contain
as elements any of the following:
%
\begin{itemize}
\item Two-component lists of the form \code{\{parameter, value\}},
  which indicates that the \code{parameter} is set to \code{value} at
  the defined scale.  If the \code{value} should be read from the SLHA
  input file, it must be written as \code{LHInput[value]}.  Example:
  %
  \begin{lstlisting}
SUSYScaleInput = {
   {mHd2, m0^2},
   {mHu2, LHInput[mHu2]}
};
  \end{lstlisting}
  %
  In this example the parameter \code{mHd2} is set to the value of
  \code{m0^2}, and \code{mHu2} is set to the value given in the SLHA
  input file in block \code{MSOFTIN}, entry 22.  The SLHA block names
  and keys for the MSSM and NMSSM are defined in SARAH's
  \code{parameters.m} file, see the SARAH manual or
  \cite{Staub:2010jh}.  The Standard Model Yukawa couplings \code{Yu},
  \code{Yd}, \code{Ye} can be calculated automatically from the known
  Standard Model quark and lepton masses, see
  Section~\ref{sec:calculation-of-yukawa-couplings}.  To enable the
  automatic calculation, \code{value} must be set to \code{Automatic},
  see the example above.

@@

%\cite{Staub:2008uz}
\bibitem{Staub:2008uz}
  F.~Staub,
  %``Sarah,''
  arXiv:0806.0538 [hep-ph].
  %%CITATION = ARXIV:0806.0538;%%
  %82 citations counted in INSPIRE as of 03 Jun 2014
